sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-08-04 doc/paper-jss/paper-jss.Rmd
In progress revamp of examples Requires very latest sdmTMB

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{pdflscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde\ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

@@

```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
library("sdmTMB")
# simplify for paper:
pcod_original <- pcod # save it for later
pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
  tibble::as_tibble()
```

```{r sdmTMB-lib, eval=TRUE, echo=TRUE}
library("sdmTMB")
head(pcod, n = 3)
```

@@

```{r pcod-utms-eval, echo=TRUE, eval=TRUE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
```

@@

```{r pcod-head2, echo=TRUE}
head(pcod, n = 3)
```

@@

```{r, echo=TRUE, eval=TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

@@

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = "on"
)
```

@@

```{r pcod-eg1-summary,eval=TRUE, echo=TRUE}
fit
```

@@

```{r pcod-eg1-tidy-fe, eval=TRUE, echo=TRUE}
tidy(fit, conf.int = TRUE)
```

```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE}
sanity(fit)
```

@@

```{r pcod-eg1-suppress-depth2, eval=TRUE, echo=FALSE}
qcs_grid_original <- qcs_grid
qcs_grid$depth_scaled <- NULL # for simplicity in paper
qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
```

```{r qcs-head, echo=TRUE}
head(qcs_grid, n = 3)
```

```{r pcod-eg1-predict, echo=TRUE, eval=TRUE, cache=TRUE}
p <- predict(fit, newdata = qcs_grid)
```

@@

```{r pcod-p-tibble}
p <- tibble::as_tibble(p)
```

```{r pcod-head-predict, echo=TRUE}
head(p, n = 3)
```

@@

```{r time1, cache=TRUE}
t1 <- Sys.time()
```

```{r fit-tv, eval=TRUE, echo=TRUE, cache=TRUE, results='hide'}
fit_spatiotemporal <- sdmTMB(
  density ~ 0 + s(depth, k = 5),
  time_varying = ~ 1,
  time_varying_type = "rw",
  family = tweedie(link = "log"),
  data = pcod,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE
)
```

```{r time2, cache=TRUE}
t2 <- Sys.time()
```

```{r fit-tv-print, eval=TRUE, echo=TRUE}
fit_spatiotemporal
```

@@

```{r pcod-restore, echo=FALSE}
pcod <- pcod_original
qcs_grid <- qcs_grid_original
```

@@

```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owls <- sdmTMB(
  count ~ nao + (1 | year_factor),
  spatial_varying = ~ nao,
  family = nbinom2(link = "log"),
  data = snow,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid"
)
```

@@

```{r child="pcod.Rmd", eval=TRUE, cache=TRUE}
```

```{r child="owls.Rmd", eval=TRUE, cache=TRUE}
```

```{r child="inla-comparison.Rmd", eval=TRUE, cache=TRUE}
```

```{r child="timing.Rmd", eval=TRUE, cache=TRUE}
```

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{pdflscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

The process component of an \pkg{sdmTMB} model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID (independent and identically distributed) random intercepts (but not currently random slopes), time-varying effects, and spatially varying effects (Fig.\ \ref{fig:diagram}).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the \pkg{sdmTMB} \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{\textasciitilde\ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

## Illustrating the need for spatial random fields

We begin with a simple species distribution model of encounter probability of
Pacific Cod (TODO spp) from a trawl survey conducted off the west
coast of Vancouver Island, British Columbia, Canada. We will fit a logistic
regression of encounter probability with and without spatial random fields to
illustrate the importance of accounting for spatial correlation.

```{r libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
library(sdmTMB)
library(dplyr)
library(ggplot2)
```

```{r libs-extras}
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
```

```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
```

```{r}
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = dat, mesh = mesh, spatial = "on",
  family = binomial(link = "logit")
)
fit_bin <- update(fit_bin_rf, spatial = "off")
```

```{r dog-bin-summary, echo=TRUE}
summary(fit_bin_rf)
```

Describe what we see here

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals.

```{r}
# set.seed(1)
# pcod$resid_bin <- residuals(fit_bin)
# set.seed(1)
# pcod$resid_bin_rf <- residuals(fit_bin_rf)
# g1 <- ggplot(pcod, aes(X, Y, colour = resid_bin)) + geom_point() +
#   scale_color_gradient2() +
#   theme(legend.position = "bottom")
# g2 <- ggplot(pcod, aes(X, Y, colour = resid_bin_rf)) + geom_point() +
#   scale_color_gradient2() +
#   theme(legend.position = "bottom")
# cowplot::plot_grid(g1, g2, ncol = 2)
```

```{r, echo=FALSE, results='hide'}
test_autocor <- function(obj) {
  set.seed(1)
  s <- simulate(obj, nsim = 500)
  pr <- predict(obj, type = "response")$est
  r <- DHARMa::createDHARMa(
    simulatedResponse = s,
    observedResponse = pcod$present,
    fittedPredictedResponse = pr
  )
  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
}
test_autocor(fit_bin)
test_autocor(fit_bin_rf)
```

We can also see that the AIC is considerably lower:

```{r}
AIC(fit_bin_rf, fit_bin)
```

Further testing could be done with cross validation via `sdmTMB_cv()`.

```{r}
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = dat, mesh = mesh, spatial = "on",
  family = binomial(link = "logit"), k_folds = 10
)
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = dat, mesh = mesh, spatial = "off",
  family = binomial(link = "logit"), k_folds = 10
)
```

Indeed, the log likelihood of left-out data is considerably larger for the model that includes random fields:

```{r}
cv_bin_rf$sum_loglik
cv_bin_rf$sum_loglik
```

The standard errors on our fixed effects have increased:

```{r}
tidy(fit_bin_rf, conf.int = T)
tidy(fit_bin, conf.int = T)
```

We now have an effect of depth and a spatial random field, which combined in link space result in our overall prediction. Note the difference between only the fixed effects (left) and including the spatial random field (right).

```{r}
p <- predict(fit_bin_rf, newdata = qcs_grid)
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position = "bottom") +
    ggtitle(title)
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
```

# Spatiotemporal species distribution modeling with dogfish

```{r dog-head-dat, echo=TRUE}
dat <- select(dogfish, longitude, latitude, year, catch_weight, area_swept, depth)
head(dat)
```

```{r dog-utms, echo=TRUE}
dat <- add_utm_columns(dat, c("longitude", "latitude"), units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
```

```{r dog-mesh, echo=TRUE}
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh)
```

```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  share_range = TRUE,
  silent = FALSE,
  family = tweedie()
)
```

We can use `update()` to refit our model with four additional families:
delta-gamma, delta-lognormal, delta-Poisson-link-gamma, and
delta-Poisson-link-lognormal.

```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
fit_dg <- update(fit_tw, family = delta_gamma(), spatiotemporal = list("off", "iid"))
fit_dl <- update(fit_dg, family = delta_lognormal())
fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
```

We can then compare the fits via AIC:

```{r dog-aic, echo=TRUE}
AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
  mutate(delta_AIC = AIC - min(AIC)) |> 
  arrange(delta_AIC)
```

We find that the Poisson-link-lognormal model is favoured with AIC. We next
test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic and structuring the
spatiotemporal random fields as AR(1).

```{r dog-ar1, results='hide', echo=TRUE}
fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
```

```{r dog-aci2, echo=TRUE}
AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
```

We find that the anisotropic AR(1) is favoured. On inspecting the model output
with `summary()`, we notice that the spatial random field SD has become very
small once AR(1) spatiotemporal fields were added. We therefore consider
a model where these are turned off in the second component of the delta model:

```{r dog-ar1-2, echo=TRUE, results="hide"}
fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
```

```{r dog-aic3, echo=TRUE}
AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
```

```{r dog-print, echo=TRUE}
fit <- fit_dpl_ar1_only
sanity(fit)
summary(fit)
```

```{r dog-aniso, echo=TRUE}
plot_anisotropy(fit_dpl)
```

```{r dog-grid, echo=TRUE}
head(wcvi_grid, n = 3)
grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
grid$log_depth <- log(grid$depth)
head(grid, n = 3)
```

```{r dog-pred1, echo=TRUE}
pred <- predict(fit, newdata = grid, type = "response")
```

```{r dog-pred2, echo=TRUE}
names(pred)
```

```{r plot-map, echo=TRUE}
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(vars(year)) +
    coord_fixed()
}
```

```{r dog-plot1, echo=TRUE}
pred |> filter(year %in% c(2004, 2014, 2022)) |> 
  plot_map(est) +
  labs(fill = "Density") +
  scale_fill_viridis_c(trans = "log10")
```

```{r dog-plot2, echo=TRUE}
pred |> filter(year %in% 2004) |> 
  plot_map(omega_s2) +
  scale_fill_gradient2()
```

```{r dog-plot3, echo=TRUE}
pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(epsilon_st2) +
  scale_fill_gradient2()
```

```{r dog-plot4, echo=TRUE}
pred |> filter(year %in% 2004) |> 
  plot_map(est_non_rf1) +
  scale_fill_viridis_c(trans = "log10")
```

```{r dog-depth, echo=TRUE}
nd <- data.frame(
  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
  year = max(dat$year)
)
pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA, model = NA)
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  labs(x = "Depth (m)", y = "Density")
```

```{r dog-index, echo=TRUE}
pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
```

```{r dog-index-plot, echo=TRUE}
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
```

We can write the model as TODO

```{r owl-data}
snow <- readRDS(here::here("data/SNOW_data.rds"))
```

```{r owl-fit, echo=TRUE, results = "hide"}
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh_snow,
  reml = TRUE,
  silent = FALSE
)
```

```{r owl-sanity, eval=FALSE}
sanity(fit_owl)
```

```{r owl-print, echo=TRUE}
summary(fit_owl)
```

We can extract the SVC effect with TODO

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
nsim <- 200
zeta_s <- predict(fit_owl, newdata = snow, nsim = nsim, sims_var = "zeta_s")
sims <- spread_sims(fit_owl, nsim = nsim)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- apply(exp(combined), 2, median)
```

We can make a basic plot using the following code. A more elaborate version is shown in Figure \@ref(fig:owl-plot-fancy).

```{r owl-plot-basic, echo=TRUE, eval=FALSE}
ggplot(snow, aes(X, Y)) +
  geom_point(aes(colour = nao_effect)) +
  scale_colour_viridis_c()
```

Overall, we found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Fig.\ \ref{fig:owl-nao}).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
We provide an extended version of this example in Appendix \ref{app:owls}.

```{r shapes, echo=FALSE}
if (!file.exists(here::here("data/ne_10m_lakes"))) {
  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
    "download/10m/physical/ne_10m_lakes.zip")
  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
}
```

```{r shapes-read, echo=FALSE}
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
```

```{r owl-proj2}
# project to get nice axis limits 
snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
snow2 <- snow2 |> mutate(x = X, y = Y) |> 
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```

```{r owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Map of spatially varying coefficient effects of NAO on Snowy Owl count. In the west, the lower values are near 1 implying no effect, whereas in the southeast the effect becomes positive. Points represent all count locations."}
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
```

@@

```{r sdmTMB-lib0, eval=FALSE, echo=FALSE}
# library("sdmTMB")
# # simplify for paper:
pcod_original <- pcod # save it for later
# pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |>
#   tibble::as_tibble()

```

```{r sdmTMB-lib, eval=FALSE, echo=TRUE}
library("sdmTMB")
dplyr::select(pcod, year, density, depth, lon, lat) |>
  head(n = 3)
```

@@

```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
```

@@

```{r pcod-head2, echo=TRUE, eval=FALSE}
head(pcod, n = 3)
```

@@

```{r, echo=TRUE, eval=FALSE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

@@

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE, cache=TRUE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = "on"
)
```

@@

```{r pcod-eg1-summary,eval=FALSE, echo=TRUE}
fit
```

@@

```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
tidy(fit, conf.int = TRUE)
```

```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r pcod-eg1-sanity, eval=FALSE, echo=TRUE}
sanity(fit)
```

@@

```{r pcod-eg1-suppress-depth2, eval=FALSE, echo=FALSE}
qcs_grid_original <- qcs_grid
qcs_grid$depth_scaled <- NULL # for simplicity in paper
qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
```

```{r qcs-head, echo=FALSE}
head(qcs_grid, n = 3)
```

```{r pcod-eg1-predict, echo=FALSE, eval=TRUE, cache=TRUE}
p <- predict(fit, newdata = qcs_grid)
```

@@

```{r pcod-p-tibble, eval=FALSE}
p <- tibble::as_tibble(p)
```

```{r pcod-head-predict, echo=TRUE, eval=FALSE}
head(p, n = 3)
```

@@

```{r time1, cache=TRUE, eval=FALSE}
t1 <- Sys.time()
```

```{r fit-tv, eval=FALSE, echo=TRUE, cache=TRUE, results='hide'}
fit_spatiotemporal <- sdmTMB(
  density ~ 0 + s(depth, k = 5),
  time_varying = ~ 1,
  time_varying_type = "rw",
  family = tweedie(link = "log"),
  data = pcod,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE
)
```

```{r time2, cache=TRUE, eval=FALSE}
t2 <- Sys.time()
```

```{r fit-tv-print, eval=FALSE, echo=TRUE}
fit_spatiotemporal
```

@@

```{r pcod-restore, echo=FALSE, eval=FALSE}
pcod <- pcod_original
qcs_grid <- qcs_grid_original
```

@@

```{r owl-fit-old, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owls <- sdmTMB(
  count ~ nao + (1 | year_factor),
  spatial_varying = ~ nao,
  family = nbinom2(link = "log"),
  data = snow,
  mesh = mesh,
  time = "year",
  spatial = "on",
  spatiotemporal = "iid"
)
```

@@

```{r child="pcod.Rmd", eval=FALSE, cache=TRUE}
```

```{r child="owls.Rmd", eval=FALSE, cache=TRUE}
```

```{r child="inla-comparison.Rmd", eval=FALSE, cache=TRUE}
```

```{r child="timing.Rmd", eval=FALSE, cache=TRUE}
```
